---
title: "Exploratory Data Analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction
In this exploratory analysis, we are going to examine the Diabetes Health Indicators dataset. This dataset has 253,680 observations in response to the CDC's BRFSS2015 survey and includes various health indicators and demographic information, with the ultimate goal of understanding factors that contribute to diabetes. The main predictors that we will be working with to help predict if someone will have diabetes or not will be a binary variable HighBP (whether someone has high blood pressure or not), a binary variable HighCol (whether someone has high cholesterol or not), a discrete variable BMI (Body Mass Index), a binary variable Smoker (whether someone has smoked at least 100 cigarettes in their lifetime or not), a binary variable PhysActivity (whether someone has had physical activity in the last 30 days outside of their job or not), a binary variable HvyAlcoholConsump (whether someone is a heavy drinker or not so that is considered adult men having more than 14 drinks per week or adult women having more than 7 drinks per week), a binary variable Sex (male or female), and 13-level categorical variable Age (split up into 13 groups of different age ranges). Going into the process of selecting these 8 predictors from the 21 total predictors in the dataset, we wanted to minimize variables that were prone to subjectivity as much as we could and select variables that we know based off intuition would have a strong effect on a person having diabetes or not. This is why HighBP, HighCol and BMI were selected as these are factors known to strongly contribute to diabetes. Also, smoking, not exercising and heavy alcohol consumption are also know activities that lead to a plethora of health problems which is why they are included as well. Age and sex were selected just due to curiosity to see the prevalence of diabetes between sexes and age groups as there would likely be a trend with more of one sex having diabetes and more of an age group or a set of age groups having diabetes as well (ex. older people).

The purpose of this EDA is to validate our health-related intuition that was used in selecting a subset of predictors to further look into. This could be how strong their relationships are with the response or any patterns or anomalies that could affect the predictive power of each predictor. Also, we want to check for any missing values to ensure our data is prepared for modeling. 

The ultimate goal of modeling is to first refine the 8 original predictor set if need be to improve model performance given EDA findings. Then, we will build different predictive models based off the refined feature set and pick the one that performs the best on a test set or unseen data. This will show which model can most accurately predict individuals who are at risk of diabetes based on a set of factors.

## Data
```{r}
# Load in required packages
library(readr)
library(dplyr)
# Load in dataset
dat <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")
# Checking for missing values
sum(is.na(dat))
# Factor conversion for response and selected predictors that are categorical
dat$Diabetes_binary <- factor(dat$Diabetes_binary, levels = c(0, 1), labels = c("No.Diabetes", "Diabetes"))
dat$HighBP <- factor(dat$HighBP, levels = c(0, 1), labels = c("No High BP", "High BP"))
dat$HighChol <- factor(dat$HighChol, levels = c(0, 1), labels = c("No High Cholesterol", "High Cholesterol"))
dat$Smoker <- factor(dat$Smoker, levels = c(0, 1), labels = c("Non-Smoker", "Smoker"))
dat$HvyAlcoholConsump <- factor(dat$HvyAlcoholConsump, levels = c(0, 1), labels = c("Non-Heavy Drinker", "Heavy Drinker"))
dat$PhysActivity <- factor(dat$PhysActivity, levels = c(0, 1), labels = c("Has Not Exercised Outside of Work", "Has Exercised Outside of Work"))
dat$Sex <- factor(dat$Sex, levels = c(0, 1), labels = c("Female", "Male"))
dat$Age <- factor(dat$Age, 
                       levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13), 
                       labels = c("18-24", "25-29", "30-34", "35-39", "40-44", 
                                  "45-49", "50-54", "55-59", "60-64", "65-69", 
                                  "70-74", "75-79", "80 or older"))
#Rename response variable for simplicity
dat <- dat |> rename(Diabetes = Diabetes_binary)
cols_to_keep <- c("Diabetes", "HighBP", "HighChol", "BMI", "Smoker", "PhysActivity", "HvyAlcoholConsump", "Sex", "Age")
dat <- dat[,cols_to_keep]
```


## Summarizations

To take a closer look at our predictors and their relationships with the response variable, we created numerical and visual summaries for each predictor. Looking at the numerical summaries (contingency tables) of the categorical predictors, we see that about 3/4 of the people observed here that have diabetes also have high blood pressure. This ratio though is not as strong when looking at high cholesterol as only 2/3 of subjects that have diabetes also have high cholesterol which is interesting to note as all patients that have diabetes may have only high blood pressure, and not always both. Looking at smoking, only about half of subjects with diabetes smoke as well. This is interesting as smoking may not have as big of an influence as we first thought. Moving onto PhysActivity, only about 37% of people with diabetes have not exercised out of work. This variable is a bit misleading though as it says exercises at least once in the past 30 days outside of work. That means subjects could have only exercised once in 30 days but that would still count for this variable which really hurts the value in the usage of this predictor. Looking at HvyAlcoholConsump, less than 5% of the people with diabetes are considered heavy drinkers, so this pretty much eliminates any usage of this predictor in the upcoming models. The reason for extremely low value could be the extreme definition of a heavy drinker being 14 drinks for males and 7 drinkers for females per week which is quite high on a week to week basis. The more likely scenario is there are a bunch of people that drink a bit less than this, but cannot be captured by the definition of this predictor. Looking at Sex, there are a bit more females with diabetes amounting to 2000 more, but there are also almost 30000 more females in the study than males. This shows how much more common diabetes is in males than females. For age, it looks like in the younger age groups like 18-24 and 25-29, less than 2% of participants have diabetes. Moving to a middle age group like 45-49, this percentage increases to around 9%. For the older age groups like 75-79 and 80+, this percentage increases to around 20%. This shows how diabetes certainly becomes more common as you age. Looking at the numeric summary of BMI, a median BMI of 27 shows that we are dealing with a slightly overweight sample which is important to take into account as a higher BMI can be caused by or a side effect of other lifestyle factors present in this dataset. Looking at the boxplot for BMI comparing people with and without diabetes, the people with diabetes certainly have a higher average BMI than those that do not as the BMI of the 75th percentile of people without diabetes is approximately equal to the BMI of the 50th percentile of people with diabetes. This is again emphasized in the overlapping histgrams for BMI as the no diabetes groups is certainly more right skewed than the diabetes group as the individuals with no diabetes tend to have lower BMI's generally. The barplots show a visual representation of the contingency tables again remphasizing the relationships observed above. This can be very clearly seen in the age barplot as the 35-39 and 75-79 age groups have roughly the same number of people without diabetes, but the 75-79 age group has a substantially higher amount of people with diabetes. This again reinforces the fact that age plays a signficant role in determining if someone has diabetes or not especially in the extreme groups.
```{r}
# Load in required packages
library(ggplot2)
# Check structure of data
str(dat)
# Check basic numerical summaries of non-categorical variables and frequencies in each level for categorical variables
summary(dat)
# Contingency tables of each categorical predictor with the response to see how they relate and interact
# List of categorical predictor variables
predictor_vars <- c("HighBP", "HighChol", "Smoker", "PhysActivity", "HvyAlcoholConsump", "Sex", "Age")
# Loop through each variable to create and display contingency tables
for (var in predictor_vars) {
  # Create the contingency table
  cat("Table for Diabetes vs", var, "\n")
  print(table(dat$Diabetes, dat[[var]]))
  cat("\n")  # Add a blank line for readability
}
# Numerical summary of BMI
summary(dat$BMI)
# Box plot of BMI distribution for people with and w/o diabetes
ggplot(dat, aes(x = Diabetes, y = BMI, fill = Diabetes)) +
  geom_boxplot() +
  labs(title = "BMI Distribution by Diabetes Status", x = "Diabetes Status", y = "BMI")
# Overlayed Histogram as another way to compare BMI by Diabetes Status
ggplot(dat, aes(x = BMI, fill = Diabetes, color = Diabetes)) +
  geom_histogram(position = "identity", binwidth = 1, alpha = 0.5, bins = 30) +
  labs(title = "Histogram of BMI by Diabetes Status", x = "BMI", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("lightblue", "salmon")) +
  scale_color_manual(values = c("darkblue", "darkred")) +
  theme(legend.title = element_blank())
# Loop through each variable to create and display bar plots
for (var in predictor_vars) {
  # Create the bar plot
  p <- ggplot(dat, aes_string(x = var, fill = "Diabetes")) +
    geom_bar(position = "dodge", color = "black") +
    labs(title = paste("Bar Plot of", var, "by Diabetes Status"),
         x = var,
         y = "Count") +
    scale_fill_manual(values = c("lightblue", "salmon")) +
    theme_minimal() +
    theme(legend.title = element_blank())
  
  # Print the plot
  print(p)
}
```



[Click here for the Modeling Page](Modeling.html)